---
title: "From Arbitrary Charting to Precise Hourly Urine Rates and AKI Staging: A Comprehensive ICU Database Analysis"
author: "Ariel A Hasidim, Matthew A Klein, Itamar Ben Shitrit, Lior Fuchs"
date: "2023-05-07"
output: 
  html_document:
    code_folding: hide
    toc: yes
    number_sections: no
    toc_float: yes
    toc_collapsed: yes
    toc_depth: 3
    theme: simplex
    code_download: no
params: 
  billing_account: "CHANGE TO BILLING ACCOUNT"
---
<!-- ################################ NOTICE ################################### -->
<!-- # 1. Running and creating the hourly-adjusted UO and AKI tables in the    # -->
<!-- #    associated billing account/project name is a preliminary requirement.# -->
<!-- #    For more information and instruction see the GitHub repository.      # -->
<!-- #                                                                         # -->
<!-- # 2. Please use 'Knit with Parameters' to provide your billing account    # -->
<!-- #    for Google's BigQuery that is credentialed for PhysioNet/MIMIC-IV.   # -->
<!-- ########################################################################### -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

# Installs pacman ("package manager") if needed
if (!require("pacman")) install.packages("pacman")

# Use pacman to load add-on packages as desired
pacman::p_load(pacman, DBI, bigrquery, tidyverse, reshape2, finalfit, psych, ggplot2, 
               ggbreak,scales,ggpmisc, ggforce, rlang, quantreg, broom, kableExtra, 
               rmdformats, table1, flextable, survival, survminer, nortest, gtsummary,
               ggsci, ggsurvfit) 

# Establish connection with BigQuery and billing acount with the correct access
con <- dbConnect(
  bigrquery::bigquery(),
  project = "physionet-data",
  dataset = "mimiciv_icu",
  billing = params$billing_account
)

dbListTables(con)
```

```{r Quering BigQuery, cache=FALSE, include=FALSE}
raw_uo <- dbGetQuery(con, "SELECT * FROM `mimic_uo_and_aki.a_urine_output_raw`")

raw_uo_excluions_duplicates <- dbGetQuery(con, statement = read_file('sql/raw_uo_excluions_duplicates.sql'))

uo_rate_including_null_collection_period <- dbGetQuery(con, "SELECT * FROM `mimic_uo_and_aki.b_uo_rate`")
uo_rate <- uo_rate_including_null_collection_period %>%
  drop_na(TIME_INTERVAL)
uo_rate_true <- uo_rate %>% filter(VALIDITY == TRUE)
uo_rate_true_foley <- uo_rate_true %>% filter(SOURCE == "Foley")

hourly_uo <- dbGetQuery(con, "SELECT * FROM `mimic_uo_and_aki.c_hourly_uo`")
uo_ml_kg_hr <- hourly_uo %>% 
  drop_na(HOURLY_VALID_WEIGHTED_MEAN_RATE, WEIGHT_ADMIT) %>%
  mutate(ML_KG_HR = HOURLY_VALID_WEIGHTED_MEAN_RATE / WEIGHT_ADMIT)

table_1 <- dbGetQuery(con, statement = read_file('sql/tbl1.sql'))

akis <- dbGetQuery(con, statement = read_file('sql/akis with mortality.sql'))

creat_diff <- dbGetQuery(con, statement = read_file('sql/creat diff.sql'))
```
\ 

This is a full reproduction of the study's results.

\ 

**For more details see:**

-   Full article at Scientific Data: [doi]()
-   [GitHub repository]()

### Citation

.............

.............

------------------------------------------------------------------------

# Population and Study's Sample

**Count all UO records (before exclusions):**

```{r}
nrow(raw_uo)
```

\ 
**Count distinct patients:**

```{r}
n_distinct(raw_uo$SUBJECT_ID)
```

\ 
**Count distinct hospital admissions:**

```{r}
n_distinct(raw_uo$HADM_ID)
```

\ 
**Count distinct ICU stays:**

```{r}
n_distinct(raw_uo$STAY_ID)
```

\ 
**Count total ICU days of UO monitoring:**

```{r}
hourly_uo %>% 
  drop_na(HOURLY_VALID_WEIGHTED_MEAN_RATE) %>%
  nrow() / 24
```

\ 

## Figure for Raw UO Records Data
```{r}
raw_uo %>% 
  mutate(
    LABEL = case_when(
      LABEL == "GU Irrigant/Urine Volume Out" ~ "GU Irrig. Out",
      LABEL == "GU Irrigant Volume In" ~ "GU Irrig. In",
      .default = LABEL
    )
  ) %>% count(LABEL, sort = TRUE) %>%
   ggplot(aes(x=reorder(LABEL, -n), y=n)) +
   geom_bar(stat="identity") +
   xlab("") +
   ylab("") +
   geom_bar(stat="identity", fill="steelblue") +
   geom_text(aes(label=n), vjust=-0.6, color="black", size=3) +
   theme_minimal() +
   theme(axis.text.y=element_blank()) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
     labs(
         title = "Frequency of Urine Output Charting by Source"
       ) +
       theme(
         plot.title = element_text(color = "#0099F8", size = 16, face = "bold"),
         plot.subtitle = element_text(size = 10, face = "bold"),
         plot.caption = element_text(face = "italic")
       )
```

\ 

## Duplication Analysis

### Check for Distinctiveness

First, we are basing distinctive rows in the raw UO data.

**Count distinct raw rows:**

```{r}
raw_uo %>% n_distinct()
```

Conclusion: the original raw query does not have duplicates (all rows are distinct by all columns)

\ 

### Check for Simultaneous Charting

```{r}
raw_uo_excluions_duplicates$same_value <- as.factor(raw_uo_excluions_duplicates$same_value)
raw_uo_excluions_duplicates$label <- as.factor(raw_uo_excluions_duplicates$label)
raw_uo_excluions_duplicates$label <- factor(raw_uo_excluions_duplicates$label, levels = as.factor(names(sort(table(raw_uo_excluions_duplicates$label),
                                  decreasing = TRUE))))
label(raw_uo_excluions_duplicates$same_value) <- "Equal values"
label(raw_uo_excluions_duplicates$label) <- "Sources"
table1(~label | same_value,  data=raw_uo_excluions_duplicates, overall=TRUE, droplevels=TRUE
       ) %>% t1flex(.)
```

|       Show full SQL query -----\>

```{sql, class.source = 'fold-hide', eval=FALSE, code = readLines("sql/raw_uo_excluions_duplicates.sql")}
```

In conclusion, most of the records have different values, and thus human error in duplicate record-keeping is not likely.

\ 

## Exclusion

Used for *Figure 2. Overview of study's process.*

**ICU type exclusion:**

```{r}
dbGetQuery(con, statement = read_file('sql/service_type_exclusion.sql')) %>%
  flextable()
```

|       Show full SQL query -----\>

```{sql, class.source = 'fold-hide', eval=FALSE, code = readLines("sql/service_type_exclusion.sql")}
```

\ 

**GU irrigation exclusion:**

```{r}
dbGetQuery(con, statement = read_file('sql/gu_irig_exclusion.sql')) %>%
  flextable()
```

|       Show full SQL query -----\>

```{sql, class.source = 'fold-hide', eval=FALSE, code = readLines("sql/gu_irig_exclusion.sql")}
```

\ 

**Uretral stent exclusion:**

```{r}
dbGetQuery(con, statement = read_file('sql/ure_stent_exclusion.sql')) %>%
  flextable()
```

|       Show full SQL query -----\>

```{sql, class.source = 'fold-hide', eval=FALSE, code = readLines("sql/ure_stent_exclusion.sql")}
```

\ 

**Not passing sanity check:**

```{r}
dbGetQuery(con, statement = read_file('sql/sanity.sql')) %>%
  flextable()
```

|       Show full SQL query -----\>

```{sql, class.source = 'fold-hide', eval=FALSE, code = readLines("sql/sanity.sql")}
```

\ 

**Total raw urine output after exclusion ("eligible records, before collection time"):**

```{r}
dbGetQuery(con, statement = read_file('sql/raw eligible.sql')) %>%
  flextable()
```

|       Show full SQL query -----\>

```{sql, class.source = 'fold-hide', eval=FALSE, code = readLines("sql/raw eligible.sql")}
```

\ 

**Exclusion of first volume in each compartment per ICU stay:**

```{r}
uo_rate_including_null_collection_period %>%
  filter(is.na(TIME_INTERVAL)) %>%
  nrow()
```

\ 

**Proportion of valid hours covered:**

```{r}
nrow(hourly_uo %>%
       filter(!is.na(HOURLY_VALID_WEIGHTED_MEAN_RATE))) / nrow(hourly_uo)
```

\ 

**Count UO records by anatomical compartment:**

```{r}
uo_rate %>% 
  mutate(agg_group = case_when(SOURCE == "Foley" |
                                 SOURCE == "Condom Cath" |
                                 SOURCE == "Straight Cath" |
                                 SOURCE == "Suprapubic" |
                                 SOURCE == "Void" ~ "Urinary bladder",
                               TRUE ~ SOURCE)
  ) %>%
           group_by(agg_group) %>%
   dplyr::summarise(N = n()
  ) %>% 
  arrange(desc(N)) %>% 
  flextable()
```

\ 

## Table 1 - Patient's characteristics

```{r}
table_1$SERVICE <- as.factor(table_1$SERVICE)
table_1$admission_age <- as.numeric(table_1$admission_age)
table_1$weight_admit <- as.numeric(table_1$weight_admit)
table_1$height_first <- as.numeric(table_1$height_first)
table_1$creat_first <- as.numeric(table_1$creat_first)

table_1 <- table_1 %>%
  mutate(
    race =
      case_when(
        grepl("asian", race, ignore.case = TRUE) ~ "Asian",
        grepl("black", race, ignore.case = TRUE) ~ "African American",
        grepl("white", race, ignore.case = TRUE) ~ "Caucasian",
        grepl("hispanic", race, ignore.case = TRUE) ~ "Hispanic",
        grepl("other", race, ignore.case = TRUE) ~ "Other",
        grepl("native", race, ignore.case = TRUE) ~ "Other",
        grepl("MULTIPLE", race, ignore.case = TRUE) ~ "Other",
        grepl("PORTUGUESE", race, ignore.case = TRUE) ~ "Other",
        grepl("SOUTH AMERICAN", race, ignore.case = TRUE) ~ "Other",
        TRUE ~ as.character(NA)
      )
  )

uo_for_table1 <- uo_rate_including_null_collection_period %>%
  group_by(STAY_ID) %>%
  summarise(
    count = n(),
    volumes = mean(VALUE, na.rm = TRUE),
    collection_times = mean(TIME_INTERVAL, na.rm = TRUE),
    rates = mean(HOURLY_RATE, na.rm = TRUE),
    validity = mean(if_else(VALIDITY == TRUE, 1, 0)),
    ml_kg_hr = mean(HOURLY_RATE / WEIGHT_ADMIT, na.rm = TRUE)
  )

t1a <- table_1 %>%
  select(
    admission_age,
    gender,
    weight_admit,
    charlson_comorbidity_index,
    sofa_first_day,
    ckd,
    creat_first,
    race,
    icu_days,
    hospital_days,
    rrt_binary,
    hospital_expire_flag
  ) %>%
  tbl_summary(
    type = list(
      c(hospital_expire_flag,
        ckd,
        rrt_binary) ~ "dichotomous",
      c(admission_age,
        weight_admit,
        creat_first) ~ "continuous"
    ),
    statistic = c(admission_age,
                  weight_admit,
                  creat_first) ~ "{mean} ({sd})",
    missing = "no",
    label = list(
      admission_age ~ "Age",
      gender ~ "Gender",
      weight_admit ~ "Weight",
      charlson_comorbidity_index ~ "CCI",
      sofa_first_day ~ "Admission SOFA",
      ckd ~ "CKD, Stage 1-4",
      creat_first ~ "First Creatinine",
      race ~ "Ethnicity",
      icu_days ~ "Total days in ICU",
      hospital_days ~ "Total days in hospital",
      rrt_binary ~ "Renal replacement therapy",
      hospital_expire_flag ~ "Mortality"
    )
  ) %>%
  add_n() %>%
  bold_labels()

display_prec <- function(x) mean(x)*100

t1b <- uo_for_table1 %>%
  select(count,
         volumes,
         collection_times,
         validity,
         rates,
         ml_kg_hr) %>%
  tbl_summary(
    type = list(
      c(volumes,
        collection_times,
        rates,
        ml_kg_hr,
        validity) ~ "continuous"
    ),
    statistic = list(
      c(
        volumes,
        collection_times,
        rates,
        ml_kg_hr
      ) ~ "{mean} ({sd})",
      validity ~ "{display_prec}%"
    ),
    missing = "no",
    label = list(
      count ~ "UO Charted",
      volumes ~ "Volumes, mean",
      collection_times ~ "Collection Times, mean",
      validity ~ "Validity, %",
      rates ~ "Rates, mean",
      ml_kg_hr ~ "Rate to Weight, mean"
    )
  ) %>%
  add_n() %>%
  bold_labels()

tbl_stack(list(t1a, t1b),
          group_header = c("ICU Stay", "UO Charting"))
```

\ 

## Table 2 - UO records characteristics

```{r}
uo_rate$SOURCE <- as.factor(uo_rate$SOURCE)
uo_rate$VALIDITY <- as.factor(uo_rate$VALIDITY)
uo_rate$SERVICE <- as.factor(uo_rate$SERVICE)

table1(~VALUE + TIME_INTERVAL + SOURCE + VALIDITY| SERVICE,  data=uo_rate, overall=TRUE, droplevels=TRUE
       ) %>% t1flex(.)
```

\ 

## Data for single patient example

The data that was used for single patient sample in table 1 and figure 6.

### Raw UO records:

```{r}
raw_uo_as_character <- raw_uo %>%
  filter(STAY_ID == 36871275)
raw_uo_as_character[] <- lapply(raw_uo_as_character, as.character)

raw_uo_as_character %>%
  select(-SUBJECT_ID, -HADM_ID, -STAY_ID, -SERVICE) %>%
  arrange(., CHARTTIME) %>%
  slice_head(n=15) %>%
  flextable()
```

\ 

### UO Rates
```{r}
uo_rate %>%
  filter(STAY_ID == 36871275) %>%
  select(-HADM_ID, -STAY_ID, -WEIGHT_ADMIT, -SERVICE) %>%
  arrange(., CHARTTIME) %>%
  slice_head(n=20) %>%
  flextable()
```

\ 

### Hourly-Adjusted UO
```{r}
hourly_uo %>%
  filter(STAY_ID == 36871275) %>%
  select(-STAY_ID, -WEIGHT_ADMIT) %>%
  arrange(., T_PLUS) %>%
  slice_head(n=20) %>%
  flextable()
```

------------------------------------------------------------------------

# Raw data analysis

## Collection Periods

```{r}
uo_rate %>% group_by(SOURCE) %>%
   dplyr::summarise(N = n(),
                   Mean = round(mean(TIME_INTERVAL),0),
                   SD = round(sd(TIME_INTERVAL),0),
                   '25th' = round(quantile(TIME_INTERVAL, 0.25),0),
                   '50th' = round(quantile(TIME_INTERVAL, 0.50),0),
                   '75th' = round(quantile(TIME_INTERVAL, 0.75),0),
                   '95th' = round(quantile(TIME_INTERVAL, 0.95),0),
                   # Min = round(min(TIME_INTERVAL),0),
                   Max = round(max(TIME_INTERVAL),0)
  ) %>% 
  arrange(desc(N)) %>% 
  flextable()
```

```{r}
ggplot(data = uo_rate, aes(x = TIME_INTERVAL / 60)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~SOURCE, scales = "free") +
  xlim(-1, 20) +
  labs(
          title = "Collection Periods",
          x = "Time interval (hr)",
          y = "Frequency"
        ) +
        theme(
          plot.title = element_text(color = "#0099F8", size = 16, face = "bold"),
          plot.subtitle = element_text(size = 10, face = "bold"),
          plot.caption = element_text(face = "italic")
        )
```

\ 

## Volumes and Collection Periods

```{r}
uo_rate %>% group_by(SOURCE) %>%
  dplyr::summarise(N = n(),
                   Mean = round(mean(VALUE),0),
                   SD = round(sd(VALUE),0),
                   '25th' = round(quantile(VALUE, 0.25),0),
                   '50th' = round(quantile(VALUE, 0.50),0),
                   '75th' = round(quantile(VALUE, 0.75),0)
  ) %>% 
  arrange(desc(N)) %>% 
  flextable()
```

```{r}
ggplot(data = uo_rate, aes(x = VALUE)) +
  facet_wrap(~SOURCE, scales = "free") +
  geom_histogram(binwidth = 50) +
  xlim(-25, 1100) +
  labs(
        title = "Volumes",
        x = "Volume (ml)",
        y = "Frequency"
      ) +
      theme(
        plot.title = element_text(color = "#0099F8", size = 16, face = "bold"),
        plot.subtitle = element_text(size = 10, face = "bold"),
        plot.caption = element_text(face = "italic")
      )
```

\ 

### Records of zero volume

**The proportion of zero value UO measurements:**

```{r}
uo_rate_count <- uo_rate %>% 
  count(SOURCE, sort = TRUE)

uo_rate_0_count <- uo_rate %>% 
  filter(VALUE == 0) %>% 
  count(SOURCE, sort = TRUE)

count_uo_zero_vs_all <- left_join(uo_rate_count, 
                                  uo_rate_0_count, by = "SOURCE") %>% 
  mutate(PROPORTION = n.y / n.x) %>%
  pivot_longer(cols = n.y:n.x, names_to = "type")
  
count_uo_zero_vs_all %>%
  ggplot(aes(x=reorder(SOURCE, -value), y=value, fill=type)) +
    geom_bar(position="fill", stat="identity") +
    xlab("") +
    ylab("") +
    scale_fill_brewer(palette="Paired") +  
    geom_text(aes(label=ifelse(type == "n.y", paste((round(PROPORTION, 3) * 100), "%"), "")), 
          color="black", 
          size=3.5, 
          vjust=-1,
          position="fill") +
    theme_minimal() +
    theme(axis.text.y=element_blank()) +
    theme(legend.position="none") +
  theme(axis.text.y=element_blank()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
     labs(
         title = "Proportion of zero value raw output count",
       ) +
       theme(
         plot.title = element_text(color = "#0099F8", size = 16, face = "bold"),
         plot.subtitle = element_text(size = 10, face = "bold"),
         plot.caption = element_text(face = "italic")
       )
```{r}
uo_rate_0 <- uo_rate %>% filter(VALUE == 0) 
S2_d <- uo_rate_0 %>% group_by(SOURCE) %>%
  dplyr::summarise(N = n(),
                   Mean = round(mean(TIME_INTERVAL),0),
                   SD = round(sd(TIME_INTERVAL),0),
                   '25th' = round(quantile(TIME_INTERVAL, 0.25),0),
                   '50th' = round(quantile(TIME_INTERVAL, 0.50),0),
                   '75th' = round(quantile(TIME_INTERVAL, 0.75),0),
                   '95th' = round(quantile(TIME_INTERVAL, 0.95),0),
                   Min = round(min(TIME_INTERVAL),0),
                   Max = round(max(TIME_INTERVAL),0)
  ) %>% 
  arrange(desc(N)) %>% 
  gt() %>%
  fmt_number(use_seps = TRUE, decimals = 0)

S2_d
```

------------------------------------------------------------------------

# Adjusting for hourly UO

## UO Rate

```{r}
uo_rate %>% group_by(SOURCE) %>%
    dplyr::summarise(N = n(),
                   Mean = round(mean(HOURLY_RATE),0),
                   SD = round(sd(HOURLY_RATE),0),
                   '5th' = round(quantile(HOURLY_RATE, 0.05),0),
                   '10th' = round(quantile(HOURLY_RATE, 0.1),0),
                   '25th' = round(quantile(HOURLY_RATE, 0.25),0),
                   '50th' = round(quantile(HOURLY_RATE, 0.50),0),
                   '75th' = round(quantile(HOURLY_RATE, 0.75),0)
                   # ,
                   # '95th' = round(quantile(HOURLY_RATE, 0.95),0),
                   # Min = round(min(HOURLY_RATE),0),
                   # Max = round(max(HOURLY_RATE),0)
  ) %>% 
  arrange(desc(N)) %>% 
  flextable()
```

```{r}
ggplot(data = uo_rate, aes(x = HOURLY_RATE)) +
  geom_histogram(binwidth = 20) +
  facet_wrap(~SOURCE, scales = "free") +
  xlim(-10, 500) +
  labs(
        title = "UO Rates",
        subtitle = "by source",
        x = "Rate (ml/hr)",
        y = "Frequency"
      ) +
      theme(
        plot.title = element_text(color = "#0099F8", size = 16, face = "bold"),
        plot.subtitle = element_text(size = 10, face = "bold"),
        plot.caption = element_text(face = "italic")
      )
```

\ 

### Low UO Rate Analysis

```{r}
uo_rate %>%
  filter(SOURCE == "Foley",
         HOURLY_RATE < 500) %>%
  # arrange(STAY_ID) %>%
  # slice_head(n = 1000000) %>%
ggplot(aes(x = HOURLY_RATE, y = TIME_INTERVAL)) +
  geom_smooth(alpha = 0.1, linewidth = 1) + 
  geom_hline(yintercept=60, size = 0.3, color = "#cccccc") +
  geom_vline(xintercept=20, size = 0.3, color = "black", linetype = "dashed") +
  scale_x_continuous(breaks = c(0, 20, 50, 100, 200, 300)) +
  scale_y_continuous(breaks = c(0, 60, 100, 200)) +
  coord_cartesian(xlim = c(-10, 350), ylim = c(0,200)) +
  labs(x="Urine output rate (ml/hr)", y = "Time interval (min)",
       title="The Association Between UO Rates and Collection Periods",
       subtitle = "Smoothed conditional means for records of Foley catheter") +
  theme_classic() + # remove panel background and gridlines
  theme(
    plot.title = element_text(color = "#0099F8", size = 16, face = "bold"),
    plot.subtitle = element_text(size = 10, face = "bold"),
    plot.caption = element_text(face = "italic")
  )
```

```{r, cache=TRUE}
uo_rate_qreg <- uo_rate %>%
  filter(SOURCE == "Foley") %>%
  arrange(STAY_ID) %>%
  slice_head(n = 200000)

#### OLS
# lm <- lm(data=uo_rate_qreg, 
#          formula =  TIME_INTERVAL ~ 
#            HOURLY_RATE)
# 
# ols <- as.data.frame(coef(lm))
# ols.ci <- as.data.frame(confint(lm, level = 0.95)) 
# ols2 <- cbind(ols, ols.ci)
# ols2 <- tibble::rownames_to_column(ols2, var="term")

#### Quantile
rq(TIME_INTERVAL ~ 
  HOURLY_RATE, seq(0.20, 0.80, by = 0.10), data=uo_rate_qreg) %>%
  tidy(se.type = "rank", conf.int = TRUE, conf.level = 0.95) %>%
  filter(!grepl("factor", term)) %>%
ggplot(aes(x=tau,y=estimate)) +
  scale_y_continuous(limits = symmetric_limits) +
  ##### quantilie results
  # geom_point(color="#27408b", size = 0.3) + 
  # geom_ribbon(aes(ymin=conf.low,ymax=conf.high),alpha=0.25, fill="#555555")+
  geom_line(color="black", linetype = "dotdash", size = 0.3) + 
  geom_point(size=3, shape=21, fill="white", size = 0.3) + 
  geom_errorbar(aes(ymin=conf.low, ymax=conf.high), width=.1) +
  facet_wrap(~term, scales="free", ncol=2,
             # labeller = labeller(.cols = label_both)
             labeller = labeller(term =
                                   c("(Intercept)" = "(Intercept)",
                                     "HOURLY_RATE" = "UO Rate (ml/hr)"))
             ) +
  ##### OLS results
  # geom_hline(data = ols2, aes(yintercept= `coef(lm)`), lty=1, color="red", size=0.3)+
  # geom_hline(data = ols2, aes(yintercept= `2.5 %`), lty=2, color="red", size=0.3)+
  # geom_hline(data = ols2, aes(yintercept= `97.5 %`), lty=2, color="red", size=0.3)+
  #### Lines
  geom_hline(yintercept = 0, size=0.3) +
  labs(title="Quantile Analysis for Collection Periods as a Function of Rates") +
  theme_classic() + # remove panel background and gridlines
  theme(
    plot.title = element_text(color = "#0099F8", size = 16, face = "bold"),
    plot.subtitle = element_text(size = 10, face = "bold"),
    plot.caption = element_text(face = "italic"),
    strip.background = element_blank()
  )
```

\ 

**Collection periods for UO rate 20ml/hr or below**

```{r}
uo_rate %>% 
  filter(HOURLY_RATE <= 20) %>%
  group_by(SOURCE) %>%
   dplyr::summarise(N = n(),
                   Mean = round(mean(TIME_INTERVAL),0),
                   SD = round(sd(TIME_INTERVAL),0),
                   '25th' = round(quantile(TIME_INTERVAL, 0.25),0),
                   '50th' = round(quantile(TIME_INTERVAL, 0.50),0),
                   '75th' = round(quantile(TIME_INTERVAL, 0.75),0),
                   '95th' = round(quantile(TIME_INTERVAL, 0.95),0),
                   # Min = round(min(TIME_INTERVAL),0),
                   Max = round(max(TIME_INTERVAL),0)
  ) %>% 
  arrange(desc(N)) %>% 
  flextable()
```

```{r}
uo_rate %>% 
  filter(HOURLY_RATE <= 20) %>%
ggplot(aes(x = TIME_INTERVAL / 60)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~SOURCE, scales = "free") +
  xlim(-1, 20) +
  labs(
          title = "Collection periods for UO rate 20ml/hr or below",
          subtitle = "by source",
          x = "Time interval (hr)",
          y = "Frequency"
        ) +
        theme(
          plot.title = element_text(color = "#0099F8", size = 16, face = "bold"),
          plot.subtitle = element_text(size = 10, face = "bold"),
          plot.caption = element_text(face = "italic")
        )
```

\ 

### Mean Rate
**Mean UO rate weighted by tyme and grouped by source:**
```{r}
uo_rate_true %>% 
  group_by(SOURCE) %>%
  summarise(weighted_mean_rate = weighted.mean(HOURLY_RATE, TIME_INTERVAL)) %>%
  flextable
```

\ 

## Hourly-adjusted UO

```{r}
hourly_uo %>% drop_na(HOURLY_VALID_WEIGHTED_MEAN_RATE) %>%
    dplyr::summarise(N = n(),
                   Mean = round(mean(HOURLY_VALID_WEIGHTED_MEAN_RATE),0),
                   SD = round(sd(HOURLY_VALID_WEIGHTED_MEAN_RATE),0),
                   '25th' = round(quantile(HOURLY_VALID_WEIGHTED_MEAN_RATE, 0.25),0),
                   '50th' = round(quantile(HOURLY_VALID_WEIGHTED_MEAN_RATE, 0.50),0),
                   '75th' = round(quantile(HOURLY_VALID_WEIGHTED_MEAN_RATE, 0.75),0),
                   '95th' = round(quantile(HOURLY_VALID_WEIGHTED_MEAN_RATE, 0.95),0),
                   Min = round(min(HOURLY_VALID_WEIGHTED_MEAN_RATE),0),
                   Max = round(max(HOURLY_VALID_WEIGHTED_MEAN_RATE),0)
  ) %>% 
  arrange(desc(N)) %>% 
  flextable()
```

```{r}
hourly_uo %>% drop_na(HOURLY_VALID_WEIGHTED_MEAN_RATE) %>%
ggplot(aes(x = HOURLY_VALID_WEIGHTED_MEAN_RATE)) +
  geom_histogram(binwidth = 20) +
  xlim(-10, 500) + 
  labs(
        title = "Hourly-Adjusted UO",
        x = "Hourly UO (ml)",
        y = "Frequency"
      ) +
      theme(
        plot.title = element_text(color = "#0099F8", size = 16, face = "bold"),
        plot.subtitle = element_text(size = 10, face = "bold"),
        plot.caption = element_text(face = "italic")
      )
```

\ 

### Simple Sum Comparison
**Showing proportion of hours with less than  100ml difference):**

```{r}
adj_uo_diff <- hourly_uo %>%
  select(HOURLY_VALID_WEIGHTED_MEAN_RATE, SIMPLE_SUM) %>%
  filter(!is.na(HOURLY_VALID_WEIGHTED_MEAN_RATE)) %>%
  mutate(hourly_diff = abs(HOURLY_VALID_WEIGHTED_MEAN_RATE - SIMPLE_SUM)) %>%
  mutate(
    cutoff_10 = if_else(hourly_diff < 10, 1, 0),
    cutoff_50 = if_else(hourly_diff < 50, 1, 0),
    cutoff_100 = if_else(hourly_diff < 100, 1, 0),
    cutoff_150 = if_else(hourly_diff < 150, 1, 0),
    cutoff_200 = if_else(hourly_diff < 200, 1, 0)
  )

my_order <- c("<10", "<50", "<100", "<150", "<200")

adj_uo_diff %>%
  select(cutoff_10,
         cutoff_50,
         cutoff_100,
         cutoff_150,
         cutoff_200) %>%
  pivot_longer(cols = contains("cutoff")) %>%
  transmute(name = case_when(
    name == "cutoff_10" ~ "<10",
    name == "cutoff_50" ~ "<50",
    name == "cutoff_100" ~ "<100",
    name == "cutoff_150" ~ "<150",
    name == "cutoff_200" ~ "<200"
  ),
  value) %>%
  group_by(name) %>%
  summarise(agreement = paste0(round(mean(value) * 100, 1), "%"),
            non_agreement = paste0(round((1 - mean(
              value
            )) * 100, 1), "%")) %>%
  arrange(match(name, my_order)) %>%
  gt() %>%
  tab_header(
    title = md("**Comparison of Hourly-Adjusted UO and Simple Summation**"),
  ) %>%
  cols_label(
    name = "Cut-off (ml)",
    agreement = "Portion of Agreement",
    non_agreement = "Portion of Disagreement"
  ) %>%
  cols_align(
    align = "center"
  )

adj_uo_diff <- hourly_uo %>%
  select(HOURLY_VALID_WEIGHTED_MEAN_RATE, SIMPLE_SUM) %>%
  filter(!is.na(HOURLY_VALID_WEIGHTED_MEAN_RATE)) %>%
  mutate(no_diff = 
           ifelse((is.na(HOURLY_VALID_WEIGHTED_MEAN_RATE) &
                  is.na(SIMPLE_SUM)) |
             (!is.na(HOURLY_VALID_WEIGHTED_MEAN_RATE) &
                  !is.na(SIMPLE_SUM) &
                    abs(HOURLY_VALID_WEIGHTED_MEAN_RATE-SIMPLE_SUM) < 100), 
                  1, 
                  0),
         .keep = "none")

mean(adj_uo_diff$no_diff)
```

\ 

### Hourly UO Per Kilogram

```{r}
uo_ml_kg_hr %>%
  dplyr::summarise(N = n(),
                   Mean = round(mean(ML_KG_HR),2),
                   SD = round(sd(ML_KG_HR),2),
                   '25th' = round(quantile(ML_KG_HR, 0.25),2),
                   '50th' = round(quantile(ML_KG_HR, 0.50),2),
                   '75th' = round(quantile(ML_KG_HR, 0.75),2),
                   '95th' = round(quantile(ML_KG_HR, 0.95),2),
                   Min = round(min(ML_KG_HR),2),
                   Max = round(max(ML_KG_HR),2)
  ) %>% flextable
```

```{r}
mean_log <- log(mean(uo_ml_kg_hr$ML_KG_HR))
sd_log <- log(sd(uo_ml_kg_hr$ML_KG_HR))
ggplot() + 
  xlim(-1, 10) + 
  geom_histogram(aes(x = ML_KG_HR
                     # , y =..density..
                     ), data=uo_ml_kg_hr, binwidth = 0.05) + 
  # stat_function(fun = dlnorm, args = list(meanlog = mean_log, sdlog = sd_log, log = FALSE), size=1, color='gray') +
  labs(
        title = "Hourly-Adjusted UO per Kilogram",
        x = "Hourly volume to kg (ml/hr/kg)",
        y = "Frequency"
      ) +
      theme(
        plot.title = element_text(color = "#0099F8", size = 16, face = "bold"),
        plot.subtitle = element_text(size = 10, face = "bold"),
        plot.caption = element_text(face = "italic")
      )
```

------------------------------------------------------------------------

# Performance analysis

```{r}
akis_excluded <- akis %>%
  filter(NO_START == 0,
         # AKI_TYPE == 1,
         HADM_RESOLVED_UO_AKI_PRE == 0)

akis_excluded$PEAK_UO_STAGE <- as.character(akis_excluded$PEAK_UO_STAGE)

aki_uo_analysis <- left_join(akis_excluded, uo_ml_kg_hr, by = "STAY_ID") %>%
  drop_na(AKI_START, 
          TIME_INTERVAL_FINISH, 
          PEAK_UO_STAGE) %>%
  select(AKI_ID,
         STAY_ID,
         PEAK_UO_STAGE,
         AKI_START,
         TIME_INTERVAL_FINISH,
         UO_KG = ML_KG_HR
         ) %>%
  mutate(TIME = as.double(difftime(TIME_INTERVAL_FINISH, 
                                   AKI_START, 
                                   units = c("hour")))) %>%
  filter(TIME >= -48 & TIME <= 48)

aki_creat_analysis <- left_join(akis_excluded, creat_diff %>% select(-STAY_ID), by = "HADM_ID") %>%
  select(AKI_ID,
         STAY_ID,
         PEAK_UO_STAGE,
         AKI_START,
         CHARTTIME,
         CREAT,
         SCR_BASELINE,
         CREAT_BASLINE_DIFF,
         CREAT_BASLINE_RATIO,
         CREAT_LOWEST7_DIFF,
         CREAT_LOWEST7_RATIO
         ) %>%
  mutate(AKI_TO_CREAT = as.double(difftime(CHARTTIME, 
                                   AKI_START, 
                                   units = c("mins"))) / 60) %>%
  filter(AKI_TO_CREAT >= -72 & AKI_TO_CREAT <= 72)
```

# Prevalence of AKI among ICU stays:

showing:

1. ICU stays with identified oliguric AKI events by UO analysis
2. Total eligible ICU stays
3. The proportion of 1 out for 2

```{r}
akis %>% 
  select(STAY_ID) %>% 
  n_distinct()

nrow(table_1)

akis %>% 
  select(STAY_ID) %>% 
  n_distinct() / nrow(table_1)
```

Number and proportion of ICU stays with identified oliguric AKI events by UO analysis out of total eligible ICU stays with breakdown for maximum KDIGO-UO stage: 

```{r}
icu_stays_3rd_stage <- akis %>% 
  filter(PEAK_UO_STAGE == 3) %>%
  select(STAY_ID) %>%
  distinct_all()
icu_stays_3rd_stage %>% 
  n_distinct()
icu_stays_3rd_stage %>% 
  n_distinct() / nrow(table_1)

icu_stays_2nd_stage <- akis %>% 
  filter(PEAK_UO_STAGE == 2,
         !(STAY_ID %in% icu_stays_3rd_stage$STAY_ID)) %>%
  select(STAY_ID) %>%
  distinct_all()
icu_stays_2nd_stage %>% 
  n_distinct()
icu_stays_2nd_stage %>% 
  n_distinct() / nrow(table_1)

icu_stays_1st_stage <- akis %>% 
  filter(PEAK_UO_STAGE == 1,
         !(STAY_ID %in% icu_stays_3rd_stage$STAY_ID),
         !(STAY_ID %in% icu_stays_2nd_stage$STAY_ID)) %>%
  select(STAY_ID) %>%
  distinct_all()
icu_stays_1st_stage %>% 
  n_distinct()
icu_stays_1st_stage %>% 
  n_distinct() / nrow(table_1)
```

## First Oliguric-AKI Events
Number of identified first oliguric-AKI events in hospitalization stratified by maximum stage:

```{r}
akis_excluded %>%
  select(PEAK_UO_STAGE) %>%
  mutate(PEAK_UO_STAGE = as.factor(PEAK_UO_STAGE)) %>%
  tbl_summary()
```

```{r}
aki_uo_analysis %>% 
ggplot(aes(TIME, UO_KG, color=PEAK_UO_STAGE, fill=PEAK_UO_STAGE))  + 
           # linetype=PEAK_UO_STAGE))  + 
  geom_hline(yintercept=0.3, size = 0.3, color = "#cccccc") +
  geom_hline(yintercept=0.5, size = 0.3, color = "#cccccc") +
  geom_vline(xintercept=0, size = 0.3, color = "black", linetype = "dashed") +
  stat_summary(fun = median, geom="line") +
  scale_x_continuous(breaks = seq(-24, 48, by=6)) +
  scale_y_continuous(breaks = c(0, 0.3, 0.5)) +
  coord_cartesian(xlim = c(-12, 24)) +
  # xlim(-24, 48) +
  stat_summary(fun.min = function(z) { quantile(z,0.25) },
               fun.max = function(z) { quantile(z,0.75) },
               geom="ribbon", colour = NA, alpha=0.2) +
  labs(x="Time around AKI onset (hour)", y = "Urine output (ml/kg/hr)", 
       title="Hourly Urine Output at the Onset of Oliguric AKI Events",
       color="Maximum KDIGO-UO stage", fill="Maximum KDIGO-UO stage") + 
       # linetype="Peak AKI stage") + 
  theme_classic() + # remove panel background and gridlines
  # annotate("rect", xmin = -24, xmax = 48, ymin = 0, ymax = 0.5,
  #          alpha = .1) +
  # annotate("rect", xmin = -24, xmax = 48, ymin = 0, ymax = 0.3,
  #          alpha = .1) +
  scale_colour_manual(values=c("#00a087ff", "#4dbbd5ff", "#e64b35ff")
                      # ,labels=c("1 (N=12,911)","2 (N=10,619)","3 (N=1,300)")
                      ) +
  scale_fill_manual(values=c("#00a087ff", "#4dbbd5ff", "#e64b35ff")
                    # ,labels=c("1 (N=12,911)","2 (N=10,619)","3 (N=1,300)")
                      ) +
  theme(
    plot.title = element_text(color = "#0099F8", size = 16, face = "bold"),
    plot.subtitle = element_text(size = 10, face = "bold"),
    plot.caption = element_text(face = "italic")
  )
    
```

\ 

## Survival Analysis
```{r}
km_fit <- survfit2(Surv(FIRST_AKI_TO_DEATH_OR_DISCH, HADM_DEATH_FLAG) ~ PEAK_UO_STAGE, data = akis_excluded)

# ggsurvplot(km_fit,
#           conf.int = TRUE,
#           risk.table.col = "strata", # Change risk table color by groups
#           ggtheme = theme_bw(), # Change ggplot2 theme
#           palette = c("#E7B800", "#2E9FDF", "#CCCCCC"),
#           xlim = c(0, 30)) 

km_fit %>%
ggsurvfit(linewidth = 1) +
  add_confidence_interval() +
  add_quantile() +
  scale_ggsurvfit(x_scales = list(breaks = c(0:30))) +
  coord_cartesian(xlim = c(0, 30)) +
  theme_classic() +
  labs(x="Days", y = "Survival", 
       title="30-Days Survival Analysis",
       subtitle = "Stratified by first hospitalization AKI event peak staging", color='Maximum KDIGO-UO stage', fill='Maximum KDIGO-UO stage') +
  scale_colour_manual(values=c("#00a087ff", "#4dbbd5ff", "#e64b35ff"),
                    labels=c("1","2","3")) +
  scale_fill_manual(values=c("#00a087ff", "#4dbbd5ff", "#e64b35ff"),
                    labels=c("1","2","3")) +
  theme(legend.position = "bottom",
        plot.title = element_text(color = "#0099F8", size = 16, face = "bold"),
        plot.subtitle = element_text(size = 10, face = "bold")) +
  add_pvalue(caption = "Log-rank {p.value}")
```

**Log rank for each pair:**
```{r}
survdiff(Surv(FIRST_AKI_TO_DEATH_OR_DISCH, HADM_DEATH_FLAG) ~ PEAK_UO_STAGE, 
         data = akis_excluded)

akis_excluded_non_1 <- akis_excluded %>%
  filter(PEAK_UO_STAGE != 1)
survdiff(Surv(FIRST_AKI_TO_DEATH_OR_DISCH, HADM_DEATH_FLAG) ~ PEAK_UO_STAGE, 
         data = akis_excluded_non_1)

akis_excluded_non_2 <- akis_excluded %>%
  filter(PEAK_UO_STAGE != 2)
survdiff(Surv(FIRST_AKI_TO_DEATH_OR_DISCH, HADM_DEATH_FLAG) ~ PEAK_UO_STAGE, 
         data = akis_excluded_non_2)

akis_excluded_non_3 <- akis_excluded %>%
  filter(PEAK_UO_STAGE != 3)
survdiff(Surv(FIRST_AKI_TO_DEATH_OR_DISCH, HADM_DEATH_FLAG) ~ PEAK_UO_STAGE, 
         data = akis_excluded_non_3)
```

\ 

## Serum Creatinine Analysis
```{r}
aki_creat_analysis %>%
ggplot(aes(x=CREAT_BASLINE_DIFF)) + 
    xlim(0, 5) + 
    geom_histogram(binwidth = 0.1)
```

```{r}
creat_plot_a <- aki_creat_analysis %>% 
  filter(PEAK_UO_STAGE == 3,
         AKI_TO_CREAT >= -24,
         AKI_TO_CREAT <= 48) %>%
  mutate(AKI_TO_CREAT_BIN = cut(AKI_TO_CREAT, 
                                breaks=12, 
                                ordered_result = TRUE,
                                # labels=FALSE
                                )) %>%
ggplot(aes(factor(AKI_TO_CREAT_BIN), CREAT_LOWEST7_DIFF)) +
    geom_boxplot(linetype = "dashed", outlier.shape = NA, color="brown") +
    stat_boxplot(aes(ymin = ..lower.., ymax = ..upper..), outlier.shape = NA, color="brown", fill="orange") +
    stat_boxplot(geom = "errorbar", aes(ymin = ..ymax..), color="brown") +
    stat_boxplot(geom = "errorbar", aes(ymax = ..ymin..), color="brown") +
    stat_summary(fun.y=mean, colour="darkred", geom="point", hape=18, size=2,show_guide = FALSE) +
    stat_summary(fun.y=mean, colour="black", geom="text", size=3, show_guide = FALSE, 
                   vjust=1.8, aes( label=round(..y.., digits=2))) +
    # scale_x_discrete(labels=c('-24      ','-18      ','-12      ','-6      ','0      ','6      ','12      ','18      ','24      ','30      ','36      ','42      ','48      ')) +
    labs(x="Time to AKI start (hours)", y = "Difference from basline (mg/dl)") +
    coord_cartesian(ylim = c(-0.1, 4)) +
    theme_classic() + # remove panel background and gridlines
    theme(legend.position = "bottom",
          plot.title = element_text(color = "#0099F8", size = 16, face = "bold"),
          plot.subtitle = element_text(size = 10, face = "bold"),
          # axis.text.x = element_text(margin = margin(t = 2),
          #                            hjust="1")
          )
```

```{r}
creat_plot_b <- aki_creat_analysis %>% 
  filter(PEAK_UO_STAGE == 3,
         AKI_TO_CREAT >= -24,
         AKI_TO_CREAT <= 48) %>%
  mutate(AKI_TO_CREAT_BIN = cut(AKI_TO_CREAT, 
                                breaks=12, 
                                ordered_result = TRUE)) %>%
ggplot(aes(factor(AKI_TO_CREAT_BIN), CREAT_LOWEST7_RATIO)) +
    geom_boxplot(linetype = "dashed", outlier.shape = NA, color="brown") +
    stat_boxplot(aes(ymin = ..lower.., ymax = ..upper..), outlier.shape = NA, color="brown", fill="orange") +
    stat_boxplot(geom = "errorbar", aes(ymin = ..ymax..), color="brown") +
    stat_boxplot(geom = "errorbar", aes(ymax = ..ymin..), color="brown") +
    stat_summary(fun.y=mean, colour="darkred", geom="point", hape=18, size=2,show_guide = FALSE) +
    stat_summary(fun.y=mean, colour="black", geom="text", size=3, show_guide = FALSE, 
                   vjust=1.8, aes( label=round(..y.., digits=2))) +
    labs(x="Time to AKI start (hours)", y = "Difference from basline (mg/dl)") +
    coord_cartesian(ylim = c(1.01, 4)) +
    theme_classic() + # remove panel background and gridlines
    theme(legend.position = "bottom",
          plot.title = element_text(color = "#0099F8", size = 16, face = "bold"),
          plot.subtitle = element_text(size = 10, face = "bold"))
```

```{r}
ggarrange(creat_plot_a, creat_plot_b,
                    labels = c("a", "b"),
                    ncol = 1, nrow = 2) %>%
  annotate_figure(top = text_grob("Change in Serum Creatinine Levels", 
               color = "#0099F8", face = "bold", size = 16, hjust = 0.9))
```

------------------------------------------------------------------------

# Technical Details

## R Session Info:

```{r}
sessionInfo()
```

::: {.tocify-extend-page data-unique="tocify-extend-page" style="height: 0;"}
:::
