---
title: "**From Arbitrary Charting to Precise Hourly Urine Output and AKI Analysis: A Comprehensive ICU Database Analysis**  \n     - Supplementary Materials -"
author: "Ariel Avraham Hasidim, Matthew Adam Kleina, Itamar Ben Shitrit, Lior Fuchs"
date: "2023-07-22"
output:
  pdf_document:
    toc: true
    number_sections: true
---

<!-- ####################################### NOTICE ########################################## -->
<!-- # 1. Running "run_results.Rmd" and  creating "s_data.Rda"  is a preliminary requirement.# -->
<!-- #    For more information and instruction see the GitHub repository.                    # -->
<!-- ######################################################################################### -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(flextable)
load("s_data.Rda")
```

\newpage

# S1 - Duplication analysis
First, we verified that all UO charting for each patient is distinct by its item and chart time (`all rows`: `r all_rows_count` = `distinct rows`: `r distinct_time_item_patient_rows_count`). This means that there are no duplicate records for a single patient simultaneously from the same source (without accounting for its volume). Subsequently, a duplication analysis was conducted. To do this, we compiled all the cases of UO measurements documented simultaneously for the same patient from different sources. We then checked for potential errors in record-keeping by verifying whether the values were consistent. We found that most of the records had different values, indicating that duplicate record-keeping errors were unlikely. There were 12,569 documented duplicates, out of which 518 had the same volume. 

The most common duplicate entry was GU irrigation IN and OUT, which had 5,347 records, accounting for 44.5% of the 12,569 duplicates. These GU irrigants were excluded from our research. We also observed many duplicate entries in combination with nephrostomy, Ileoconduit, and urethral stent. Based on this analysis, we concluded that there are a total of four potential physiologic compartments for the accumulation and drainage of urine output that can be measured. This means that when we assume that the output events occurred "back-to-back" to determine that the collection period is equal to the time interval from the last chart time, we must examine each compartment individually.

## Simultaneous UO charting and corresponding source count by volume equality 

```{r}
S1_a
```

\newpage
# S2 - Volume Analysis

This section aims to clarify the relationship between UO volumes and collection periods. 

## Descriptive statistics and histogram for volume measurments by source (ml)

```{r}
S2_a
```


```{r}
S2_b
```

The patterns of the volume histograms for the different UO sources are similar to the patterns identified in the collection period analysis. These patterns result from the linear relationship between the UO volume and the duration of the collection period. On average, the longer the time interval between UO measurements, the higher the UO volume is. 

Recording zero-volume measurements is crucial for maintaining proper temporal resolution, particularly in cases of oliguria and anuria. Recording zero volume UO is typical and expected when performing methodical hourly documentation by the nursing staff. However, we must consider whether adequate documentation of zero volume values was performed for sources that relied upon spontaneous voiding (e.g., void and condom catheters).

## Proportion of zero volume charting by source

```{r}
S2_c
```


As shown above, the percentage of zero values for all sources of measurement falls between 1-2.5% and is consistent across different sources, including void and condom catheters. We can observe a higher proportion of zero-value recordings with nephrostomies, which can be explained by the correlation between low UO rate and single-kidney sources. Furthermore, a higher proportion of zero values in “Suprapubic” outputs can be attributed to unsuccessful attempts to aspirate urine with a needle when urine retention is suspected.

## Descriptive statistics of collection periods for zero-volume measurments

```{r}
S2_d
```

Although the amount of urine collected is directly related to the duration of the collection period (as urine output is continuous and expressed as volume per unit of time), paradoxically, the descriptive statistics for zero-volumes indicate longer collection times than the descriptive statistics for all measurements. For instance, when measuring zero volume for a Foley catheter, the average collection time is 103 minutes, and the 95th percentile is 240 minutes, whereas the average collection time for all measurements is 83 minutes, and the 95th percentile is 180 minutes. 

\newpage
# S3 - Rate Analysis

## Descriptive statistics and histograms of valid UO rates (ml/hr) for each time interval by source

```{r}
S3_a
```

```{r}
S3_b
```

The table and figure above demonstrate the accuracy of our meticulous approach in extracting the actual UO rates from various sources. Notably, we observed a high degree of similarity in the measures of central tendency and distributions of the condom, Foley, ileoconduit, straight, suprapubic, and void catheters. This similarity can be explained by the fact that, following the adjustment of the volume-to-time interval, all of these sources represent the total UO from both kidneys. 

Moreover, the decision to separate the right and left nephrostomy sources from the others was correct, as there is a high degree of similarity between these two sources. Additionally, the measures of central tendency of these two sources are significantly smaller than the rest, which is expected as they represent the UO of a single kidney compared to two kidneys, even though the measures of central tendency are not exactly one-half.

However, the table and figure above show unweighted values, which makes it difficult to appreciate the real physiologic urine production observed by each source and compare them. “Unweighted” means that we did not account for the duration of each UO rate when calculating the statistics. For example, one hour of UO rate 0 has the same effect on the average as one minute with 1000 ml/hr does. To overcome this obstacle, we calculated the average UO rate for each measurement source and weighted it by the collection periods, as shown in Table 9 and further discussed in the paper.

```{r}
S3_c
```

\newpage
# S4 - Hourly-Adjusted Analysis

## Descriptive statistics and histogram for hourly-adjusted UO (ml/hr)

```{r}
S4_a
```

```{r}
S4_b
```

## Descriptive statistics and histogram for hourly-adjusted UO to weight (ml/hr/kg)

```{r}
S4_c
```

```{r}
S4_d
```

By examining the descriptive statistics and histogram of hourly urine out per kilogram, we can see an unimodal, smooth, and homogenous distribution with a positive skew and no significant outliers. These characteristics are consistent with valid observations that reflect the natural variability in the study population without any major indications of unusual values, data entry errors, or subpopulations with distinct characteristics.
